---
title: "ClinVar Data Processing"
author: "Garrett Cooper"
date: 11/23/2024
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

Here we are filtering for ClinVar variants that were included in our screen, particularly those that affect the coding sequence of SMARCB1. (Data obtained from ClinVar on December, 4 2024)

```{r}

library(readr)
library(dplyr)
library(yaml)
library(stringr)

#read in config file
config <- read_yaml("./config.yaml")

#We are filtering data that was obtained from ClinVar's website on December 4, 2024

clinvar_SMARCB1 <- read_tsv(file=config$clinvar_SMARCB1_raw)

```



First we need to filter for only deletions, indels, missense, and SNVs
```{r}


#I am removing copy number gain, copy number loss, duplication
clinvar_SMARCB1_filtered1 <- clinvar_SMARCB1 %>%
  filter(clinvar_SMARCB1$`Variant type` == "single nucleotide variant"|
         clinvar_SMARCB1$`Variant type` == "Deletion"|
         clinvar_SMARCB1$`Variant type` == "Indel"|
         clinvar_SMARCB1$`Variant type` == "Insertion"|
         clinvar_SMARCB1$`Variant type` == "Microsatellite")

#7 were weird with DERL3 annotation, lots of intronic, and 3' and 5' UTR variants

clinvar_SMARCB1_filtered2 <- clinvar_SMARCB1_filtered1[!grepl("DERL3", clinvar_SMARCB1_filtered1$Name) & !is.na(clinvar_SMARCB1_filtered1$`Molecular consequence`) &  !grepl("3 prime UTR variant", clinvar_SMARCB1_filtered1$`Molecular consequence`) & !grepl("5 prime UTR variant", clinvar_SMARCB1_filtered1$`Molecular consequence`) & !grepl("intron variant", clinvar_SMARCB1_filtered1$`Molecular consequence`) ,]

#There are alot of pathogenic splice variants in ClinVar but our screen does not include these so they will be filtered out... will keep for later
clinvar_splice_variants <- clinvar_SMARCB1_filtered2 %>%
  filter(clinvar_SMARCB1_filtered2$`Molecular consequence` == "splice acceptor variant"|
         clinvar_SMARCB1_filtered2$`Molecular consequence` == "splice donor variant")


clinvar_SMARCB1_filtered3 <- clinvar_SMARCB1_filtered2 %>%
  filter(clinvar_SMARCB1_filtered2$`Molecular consequence` != "splice acceptor variant",
         clinvar_SMARCB1_filtered2$`Molecular consequence` != "splice donor variant",
         clinvar_SMARCB1_filtered2$`Molecular consequence` != "splice donor variant",
         clinvar_SMARCB1_filtered2$`Molecular consequence` != "inframe_indel",
         clinvar_SMARCB1_filtered2$`Molecular consequence` != "inframe_deletion|initiator_codon_variant",
         clinvar_SMARCB1_filtered2$`Molecular consequence` != "inframe_deletion" )


clinvar_SMARCB1_NA_protein_change <- clinvar_SMARCB1_filtered3[is.na(clinvar_SMARCB1_filtered3$`Protein change`), ]

#Theres also some funky mutations that are in frame multiple amino acid deletions and multi amino acid changes (missense variant here) that are not included in the screen so we will filter them out

clinvar_SMARCB1_synonymous <- clinvar_SMARCB1_NA_protein_change %>%
  filter(clinvar_SMARCB1_NA_protein_change$`Molecular consequence` != "inframe_deletion",
         clinvar_SMARCB1_NA_protein_change$`Molecular consequence` != "inframe_indel",
         clinvar_SMARCB1_NA_protein_change$`Molecular consequence` != "missense variant")
         


clinvar_SMARCB1_with_protein_change <- clinvar_SMARCB1_filtered3[!is.na(clinvar_SMARCB1_filtered3$`Protein change`), ]


#There's some funky one we just gotta do by hand, were putting them into a new df and will merge them back in later...
clinvar_SMARCB1_exceptions <- clinvar_SMARCB1_with_protein_change[89,]


#Removing the one exception and also the two that are not in isoform v1
clinvar_SMARCB1_with_protein_change <- clinvar_SMARCB1_with_protein_change[-c(89,165,166),]




```



Clinvar has the amino acid subsititions for different isoforms which is super annoying, so here we are filtering the Protein Change column for only the v1 amino acid values.
```{r}



extract_number <- function(text) {
  # Extract the number that comes after any letter sequence (like Met, Leu, etc.)
  numbers <- gsub(".*[A-Za-z]+([0-9]+).*", "\\1", text)
  
  # Return the extracted number
  return(numbers)
}

clinvar_SMARCB1_with_protein_change$Residue <- sapply(clinvar_SMARCB1_with_protein_change$Name, extract_number)

filter_second_column <- function(extracted_number, second_column) {
  # Split the second column by commas and trim any leading or trailing spaces
  variants <- strsplit(second_column, ",")[[1]]
  variants <- trimws(variants)  # Remove leading/trailing spaces from each variant
  
  # Find values that match the extracted number
  matching_variants <- variants[grep(paste0("[^a-zA-Z]*", extracted_number, "[^a-zA-Z]*"), variants)]
  
  # Return the matching variants as a string
  return(paste(matching_variants, collapse = ","))
}

clinvar_SMARCB1_with_protein_change$`Protein change` <- mapply(filter_second_column, clinvar_SMARCB1_with_protein_change$Residue, clinvar_SMARCB1_with_protein_change$`Protein change`)


```

Now we need to filter the exceptions that didnt work in that function above. Then we combine them
```{r}

# Create a function to extract specific values by position
extract_specific_variant <- function(text, position) {
  # Split the text by commas
  variants <- strsplit(text, ",")[[1]]
  # Trim any leading or trailing whitespace
  variants <- trimws(variants)
  # Return the value at the specified position
  return(variants[position])
}

# Define the positions you want for each row
positions <- c(1)

# Apply the function row-wise, using the specified positions
clinvar_SMARCB1_exceptions$`Protein change` <- mapply(
  extract_specific_variant,
  clinvar_SMARCB1_exceptions$`Protein change`,
  positions
)



clinvar_SMARCB1_exceptions$Residue <- sapply(clinvar_SMARCB1_exceptions$`Protein change`, extract_number)


clinvar_SMARCB1_protein_change2 <- rbind(clinvar_SMARCB1_exceptions, clinvar_SMARCB1_with_protein_change)

```

Join the synonymous and nonsynonymous mutants

```{r}

clinvar_SMARCB1_synonymous$Residue <- sapply(clinvar_SMARCB1_synonymous$Name, extract_number)


#Now lets add in the synonymous substitution so we can correspond it with the MITE data.
SMARCB1_sequence <- "MMMMALSKTFGQKPVKFQLEDDGEFYMIGSEVGNYLRMFRGSLYKRYPSLWRRLATVEERKKIVASSHGKKTKPNTKDHGYTTLATSVTLLKASEVEEILDGNDEKYKAVSISTEPPTYLREQKAKRNSQWVPTLPNSSHHLDAVPCSTTINRNRMGRDKKRTFPLCFDDHDPAVIHENASQPEVLVPIRLDMEIDGQKLRDAFTWNMNEKLMTPEMFSEILCDDLDLNPLTFVPAIASAIRQQIESYPTDSILEDQSDQRVIIKLNIHVGNISLVDQFEWDMSEKENSPEKFALKLCSELGLGGEFVTTIAYSIRGQLSWHQKTYAFSENPLPTVEIAIRNTGDADQWCPLLETLTDAEMEKKIRDQDRNTRRMRRLANTAPAW"

clinvar_SMARCB1_synonymous$`Protein change` <- sapply(clinvar_SMARCB1_synonymous$Residue, function(residue) {
  # Check if Residue is not NA
  if (!is.na(residue)) {
    # Get the corresponding amino acid from the sequence
    amino_acid <- substr(SMARCB1_sequence, residue, residue)
    # Format it as "AminoAcidNumberAminoAcid"
    return(paste0(amino_acid, residue, amino_acid))
  } else {
    return(NA)  # Keep NA if the Residue is NA
  }
})


clinvar_SMARCB1_filtered4 <- rbind(clinvar_SMARCB1_synonymous, clinvar_SMARCB1_protein_change2)

#Now lets filter it for only the columns we want
clinvar_SMARCB1_filtered5 <- clinvar_SMARCB1_filtered4[,c(1,3,4,5,8,9,10,12,14,15,16,26)]

#Now we have to reformat the Protein change column so it matches exactly with the mite data

clinvar_SMARCB1_filtered5 <- clinvar_SMARCB1_filtered5 %>%
  mutate(`Protein change` = gsub("fs$", "Fs", `Protein change`), # Replace 'fs' with 'Fs'
         `Protein change` = gsub("\\*$", "X", `Protein change`)) # Replace '*' with 'X'

write_tsv(clinvar_SMARCB1_filtered5, "./data/clinvar/clinvar_SMARCB1_coding_variants.tsv")



clinvar_SMARCB1_missense <- clinvar_SMARCB1_filtered5 %>%
  filter(str_detect(`Molecular consequence`, "missense"))


pathogenic_SMARCB1 <- clinvar_SMARCB1_missense %>%
  filter(clinvar_SMARCB1_missense$`Germline classification` == "Pathogenic" | clinvar_SMARCB1_missense$`Germline classification` == "Pathogenic/Likely pathogenic" | clinvar_SMARCB1_missense$`Germline classification` == "Likely pathogenic")

VUS_SMARCB1 <- clinvar_SMARCB1_missense %>%
  filter(clinvar_SMARCB1_missense$`Germline classification` == "Uncertain significance" | clinvar_SMARCB1_missense$`Germline classification` == "Conflicting classifications of pathogenicity")

Benign_SMARCB1 <- clinvar_SMARCB1_missense %>%
  filter(clinvar_SMARCB1_missense$`Germline classification` == "Likely benign" | clinvar_SMARCB1_missense$`Germline classification` == "Benign" | clinvar_SMARCB1_missense$`Germline classification` == "Benign/Likely benign")
  
  


```








Next let's filter the BRCA1 Variants

```{r}

clinvar_BRCA1 <- read_tsv(file=config$clinvar_BRCA1_raw)


clinvar_BRCA1_missense <- clinvar_BRCA1 %>% 
  filter(str_detect(`Molecular consequence`, "missense"))


clinvar_BRCA1_missense$Residue <- sapply(clinvar_BRCA1_missense$Name, extract_number)


clinvar_BRCA1_missense$`Protein change` <- mapply(filter_second_column, clinvar_BRCA1_missense$Residue, clinvar_BRCA1_missense$`Protein change`)

clinvar_BRCA1_missense <- clinvar_BRCA1_missense %>%
  mutate(`Protein change` = str_extract(`Protein change`, "^[^,]+"))

clinvar_BRCA1_missense <- clinvar_BRCA1_missense %>% 
  filter(str_detect(`Name`, "NM_007294.4"))



# Replace 'your_dataframe' and 'your_column' with the actual names
row_index <- grep("NM_007294.4\\(BRCA1\\):c\\.2155A>G \\(p\\.Lys719Glu\\)", clinvar_BRCA1_missense$Name)

# View the row index
print(row_index)

clinvar_BRCA1_missense <- clinvar_BRCA1_missense %>%
  mutate(`Protein change` = if_else(row_number() == 3724, "K719E", `Protein change`))


clinvar_BRCA1_no_protein_change <- clinvar_BRCA1_missense %>%
  filter(is.na(`Protein change`) | `Protein change` == "")


clinvar_BRCA1_missense <- clinvar_BRCA1_missense %>%
  filter(!is.na(`Protein change`) & `Protein change` != "")



write_tsv(clinvar_BRCA1_missense, "./data/clinvar/clinvar_BRCA1_coding_variants_missense.tsv")



pathogenic_BRCA1 <- clinvar_BRCA1_missense %>%
  filter(clinvar_BRCA1_missense$`Germline classification` == "Pathogenic" | clinvar_BRCA1_missense$`Germline classification` == "Pathogenic/Likely pathogenic" | clinvar_BRCA1_missense$`Germline classification` == "Likely pathogenic")

VUS_BRCA1 <- clinvar_BRCA1_missense %>%
  filter(clinvar_BRCA1_missense$`Germline classification` == "Uncertain significance" | clinvar_BRCA1_missense$`Germline classification` == "Conflicting classifications of pathogenicity")

Benign_BRCA1 <- clinvar_BRCA1_missense %>%
  filter(clinvar_BRCA1_missense$`Germline classification` == "Likely benign" | clinvar_BRCA1_missense$`Germline classification` == "Benign" | clinvar_BRCA1_missense$`Germline classification` == "Benign/Likely benign")

```




Next letts filter the TP53 ClinVar Variants


```{r}

clinvar_TP53 <- read_tsv(file=config$clinvar_TP53_raw)


clinvar_TP53_missense <- clinvar_TP53 %>% 
  filter(str_detect(`Molecular consequence`, "missense"))


clinvar_TP53_missense$Residue <- sapply(clinvar_TP53_missense$Name, extract_number)


clinvar_TP53_missense$`Protein change` <- mapply(filter_second_column, clinvar_TP53_missense$Residue, clinvar_TP53_missense$`Protein change`)

clinvar_TP53_missense <- clinvar_TP53_missense %>%
  mutate(`Protein change` = str_extract(`Protein change`, "^[^,]+"))

clinvar_TP53_missense <- clinvar_TP53_missense %>% 
  filter(str_detect(`Name`, "NM_000546.6"))



# Replace 'your_dataframe' and 'your_column' with the actual names
row_index <- grep("NM_000546\\.6\\(TP53\\):c\\.755T>C \\(p\\.Leu252Pro\\)", clinvar_TP53_missense$Name)

# View the row index
print(row_index)

clinvar_TP53_missense <- clinvar_TP53_missense %>%
  mutate(`Protein change` = if_else(row_number() == 483, "L252P", `Protein change`))


clinvar_TP53_no_protein_change <- clinvar_TP53_missense %>%
  filter(is.na(`Protein change`) | `Protein change` == "")


clinvar_TP53_missense <- clinvar_TP53_missense %>%
  filter(!is.na(`Protein change`) & `Protein change` != "")



write_tsv(clinvar_TP53_missense, "./data/clinvar/clinvar_TP53_coding_variants_missense.tsv")


pathogenic_TP53 <- clinvar_TP53_missense %>%
  filter(clinvar_TP53_missense$`Germline classification` == "Pathogenic" | clinvar_TP53_missense$`Germline classification` == "Pathogenic/Likely pathogenic" | clinvar_TP53_missense$`Germline classification` == "Likely pathogenic")

VUS_TP53 <- clinvar_TP53_missense %>%
  filter(clinvar_TP53_missense$`Germline classification` == "Uncertain significance" | clinvar_TP53_missense$`Germline classification` == "Conflicting classifications of pathogenicity" )

Benign_TP53 <- clinvar_TP53_missense %>%
  filter(clinvar_TP53_missense$`Germline classification` == "Likely benign" | clinvar_TP53_missense$`Germline classification` == "Benign" | clinvar_TP53_missense$`Germline classification` == "Benign/Likely benign")

```



Next lets filter the PTEN ClinVar Variants


```{r}

clinvar_PTEN <- read_tsv(file=config$clinvar_PTEN_raw)


clinvar_PTEN_missense <- clinvar_PTEN %>% 
  filter(str_detect(`Molecular consequence`, "missense"))


clinvar_PTEN_missense$Residue <- sapply(clinvar_PTEN_missense$Name, extract_number)


clinvar_PTEN_missense$`Protein change` <- mapply(filter_second_column, clinvar_PTEN_missense$Residue, clinvar_PTEN_missense$`Protein change`)

clinvar_PTEN_missense <- clinvar_PTEN_missense %>%
  mutate(`Protein change` = str_extract(`Protein change`, "^[^,]+"))

clinvar_PTEN_missense <- clinvar_PTEN_missense %>% 
  filter(str_detect(`Name`, "NM_000314.8"))



clinvar_PTEN_no_protein_change <- clinvar_PTEN_missense %>%
  filter(is.na(`Protein change`) | `Protein change` == "")


clinvar_PTEN_missense <- clinvar_PTEN_missense %>%
  filter(!is.na(`Protein change`) & `Protein change` != "")



write_tsv(clinvar_PTEN_missense, "./data/clinvar/clinvar_PTEN_coding_variants_missense.tsv")


pathogenic_PTEN <- clinvar_PTEN_missense %>%
  filter(clinvar_PTEN_missense$`Germline classification` == "Pathogenic" | clinvar_PTEN_missense$`Germline classification` == "Pathogenic/Likely pathogenic" | clinvar_PTEN_missense$`Germline classification` == "Likely pathogenic")

VUS_PTEN <- clinvar_PTEN_missense %>%
  filter(clinvar_PTEN_missense$`Germline classification` == "Uncertain significance" | clinvar_PTEN_missense$`Germline classification` == "Conflicting classifications of pathogenicity" )

Benign_PTEN <- clinvar_PTEN_missense %>%
  filter(clinvar_PTEN_missense$`Germline classification` == "Likely benign" | clinvar_PTEN_missense$`Germline classification` == "Benign" | clinvar_PTEN_missense$`Germline classification` == "Benign/Likely benign")

```


