**Molecular Dynamics Simulation Guide**

**System Setup**
- **Platform:** Linux-based system running Amber v24
- **Hardware:** NVIDIA RTX A6000 with 49,140 MiB of memory
- **CUDA Version:** 12.6
**Input Structure**
**SMARCB1 PDB:** Obtained from AlphaFold Q12824
**Mutations:** WT structure mutated to W281P and I315R using PyMOL.
**Processing:** All structures (WT, W281P, and I315R) processed using Amber v24 in the same way as described below.



1. Prepare the PDB File
```{bash}
pdb4amber WT.pdb > WT.amber.pdb
```



2.	Set up the topology

Create a file called leap.in containing the following:
```{bash}
source leaprc.protein.ff19SB
source leaprc.water.tip3p
mol = loadPDB WT.amber.pdb
solvateBox mol TIP3PBOX 10.0
addIons mol Na+ 0
saveAmberParm mol system.prmtop system.inpcrd
quit
```

Run the following command to generate the topology:
```{bash}
tleap -f leap.in
```



3.	Energy Minimization (This takes a while to run)

Create a file called min.in for energy minimization:
```{bash}
Energy minimization
&cntrl
  imin=1,        ! Perform minimization
  maxcyc=5000,  ! Maximum number of cycles
  ncyc=2500,    ! Number of cycles for steepest descent
  cut=10.0,
&end
```

Run the minimization:
```{bash}
nohup pmemd.cuda -O -i min.in -o min.out -p system.prmtop -c system.inpcrd -r min.crd -ref system.inpcrd &
```



4.	Equilibration

a) NVT Equilibriation

Create a file called nvt.in for NVT equilibration:
```{bash}
NVT equilibration
&cntrl
  imin=0,        ! No minimization
  ntt=3,         ! Temperature coupling
  temp0=300.0,   ! Target temperature
  ntpr=100,      ! Print every 100 steps
  ntwx=100,      ! Write coordinates every 100 steps
  ntf=2,
  ntb=1,
  cut=10.0,
&end
```

Run the equilibration:
```{bash}
nohup pmemd.cuda -O -i nvt.in -o nvt.out -p system.prmtop -c min.crd -r nvt.crd -x nvt.mdcrd &
```

b) NPT Equilibriation

Create a file called npt.in for NPT equilibration:
```{bash}
NPT equilibration
&cntrl
  imin=0,          ! No minimization
  ntt=3,           ! Temperature coupling
  ntp=1,           ! Pressure coupling
  ntpr=100,
  ntwx=100,
  cut=10.0,
  taup=2.0,        ! Pressure relaxation time
  temp0=300.0,     ! Target temperature
  gamma_ln=1.0,    ! Friction coefficient for Langevin dynamics
&end
```

Run the equilibration:
```{bash}
nohup pmemd.cuda -O -i npt.in -o npt.out -p system.prmtop -c nvt.crd -r npt.crd -x npt.mdcrd &
```



5.	Run Molecular Dynamics Simulation

Create a file called md.in for the MD simulation:
```{bash}
Production MD simulation
&cntrl
  imin=0,
  ntt=3,
  ntp=1,
  ntpr=100,
  ntwx=100,
  cut=10.0,
  nstlim=200000000,  ! Number of steps to run
  dt=0.001,       ! Reduced time step
  temp0=300.0,
  gamma_ln=5.0,   ! Higher friction coefficient
  restraint_wt=2.0,  ! Restraint weight on solute atoms
  restraintmask=":1-385@CA"  ! Apply restraints to only alpha carbons
&end
```

Run the MD simulation:
```{bash}
nohup pmemd.cuda -O -i md.in -o md.out -p system.prmtop -c npt.crd -r md.crd -x md.mdcrd &
```



6.	Analyze Results

RMSD Analysis:
```{bash}
cpptraj -p system.prmtop -y md.mdcrd
rmsd :1-385@CA out rmsd.dat
run
```

RMSF Analysis:
```{bash}
cpptraj -p system.prmtop -y md.mdcrd
trajin md.mdcrd
atomicfluct out rmsf.dat byres :1-385
run
```



7. Convert trajectory to VMD readable format
```{bash}
cpptraj
parm system.prmtop 
trajin md.mdcrd
trajout md.dcd
run
```
To load into VMD, first load system.prmtop, then load the md.dcd file.



8. Plotting Trajectory Results

a)RMSD

```{r}
library(readr)
library(ggplot2)

WT_RMSD <- read.table("./results/molecular_dynamics_simulation/rmsd_WT_200ns.dat", header=FALSE)
colnames(WT_RMSD) <- c("Time", "RMSD")


I315R_RMSD <- read.table("./results/molecular_dynamics_simulation/rmsd_I315R_200ns.dat", header=FALSE)
colnames(I315R_RMSD) <- c("Time", "RMSD")

W281P_RMSD <- read.table("./results/molecular_dynamics_simulation/rmsd_W281P_200ns.dat", header=FALSE)
colnames(W281P_RMSD) <- c("Time", "RMSD")

WT_RMSD$Variant <- "WT"
I315R_RMSD$Variant <- "I315R"
W281P_RMSD$Variant <- "W281P"

combined_RMSD <- rbind(WT_RMSD, I315R_RMSD, W281P_RMSD)

# Reorder the columns for better readability
combined_RMSD <- combined_RMSD[, c("Time", "RMSD", "Variant")]

# View the combined dataframe
head(combined_RMSD)

#Save for later use
write_xlsx(combined_RMSD, "./results/molecular_dynamics_simulation/combined_rmsd.xlsx")




time_per_frame_ns <- 0.005

combined_RMSD$Time_ns <- combined_RMSD$Time * time_per_frame_ns

# Create the plot
rmsd <- ggplot(combined_RMSD, aes(x = Time_ns, y = RMSD, color = factor(Variant, levels = c("WT", "W281P", "I315R")))) +
  geom_line(size= 1.5) +  
  labs(title = "RMSD vs. Time",
       x = "Time (ns)",
       y = "RMSD (Å)",
       color = "Variant") +     # Legend title
  theme_minimal() +           # Use a minimal theme
  theme(legend.position = "top",
        axis.text = element_text(size = 30),
        axis.title = element_text(size = 34),
        legend.text = element_text(size = 26),
        legend.title = element_text(size = 34),
        plot.title = element_text(size = 36)) +
  scale_color_manual(values = c("WT" = "gray", "I315R" = "red", "W281P" = "blue"))  # Customize colors

rmsd



ggsave("./plots/molecular_dynamics_simulation/rmsd_200ns.tif", plot = rmsd, width = 34, height =20, units = "cm", device = tiff ,res = 600)

```

b) RMSF

```{r}

#Read in RMSF data
WT_RMSF <- read.table("./results/molecular_dynamics_simulation/rmsf_WT_200ns.dat", header=FALSE)
colnames(WT_RMSF) <- c("Residue", "RMSF")
W281P_RMSF <- read.table("./results/molecular_dynamics_simulation/rmsf_W281P_200ns.dat", header=FALSE)
colnames(W281P_RMSF) <- c("Residue", "RMSF")
I315R_RMSF <- read.table("./results/molecular_dynamics_simulation/rmsf_I315R_200ns.dat", header=FALSE)
colnames(I315R_RMSF) <- c("Residue", "RMSF")
WT_RMSF$Variant <- "WT"
I315R_RMSF$Variant <- "I315R"
W281P_RMSF$Variant <- "W281P"

combined_RMSF <- rbind(WT_RMSF, I315R_RMSF, W281P_RMSF)

# Reorder the columns for better readability
combined_RMSF <- combined_RMSF[, c("Residue", "RMSF", "Variant")]

# View the combined dataframe
head(combined_RMSF)

domain_data <- data.frame(
  Domain = c( "WHD","RPT1", "RPT2", "CCD"),
  Start = c( 10,183, 259, 335),
  End = c(110, 245, 319, 378),
  Color = c("gray70","green", "orange", "blue")  # Colors for each domain
)

# Create the plot
rmsf <- ggplot() +
  # Add rectangles for domains
  geom_rect(data = domain_data,
            aes(xmin = Start, xmax = End, ymin = -Inf, ymax = Inf, fill = Domain),
            alpha = 0.3) +  # Make rectangles slightly transparent
  # First add mutant points
  geom_point(data = subset(combined_RMSF, Variant != "WT"), 
             aes(x = Residue, y = RMSF, color = Variant), 
             size = 1) +  # Mutant points
  # Then add WT points, this ensures they are on top
  geom_point(data = subset(combined_RMSF, Variant == "WT"), 
             aes(x = Residue, y = RMSF, color = Variant), 
             size = 1) +  # WT points on top
  geom_line(data = combined_RMSF, aes(x = Residue, y = RMSF, color = Variant), 
            linewidth = 2, alpha = 0.9) + 
  labs(title = "RMSF by Residue with Domains",
       x = "Residue Number",
       y = "RMSF Value",
       color = "Variant",
       fill = "Domain") +  # Legend title for fill
  theme_minimal() +  # Use a minimal theme
  theme(legend.position = "top",
  axis.text = element_text(size = 30),
        axis.title = element_text(size = 34),
        legend.text = element_text(size = 26),
        legend.title = element_text(size = 34),
        plot.title = element_text(size = 36)) +
  scale_fill_manual(values = domain_data$Color, 
                    limits = c("WHD","RPT1", "RPT2", "CCD")) + 
  scale_color_manual(values = c("WT" = "gray", "W281P" = "blue", "I315R" = "red" ),  # Adjust colors if needed
                     limits = c("WT","W281P", "I315R")) +
  scale_x_continuous(limits = c(0, 385)) + 
  scale_y_continuous(limits = c(5, 25))  

rmsf

ggsave("./plots/molecular_dynamics_simulation/rmsf_200ns.tif", plot = rmsf, width = 34, height =20, units = "cm", device = tiff ,res = 600)
```

```



